Pottery decor as networks on the Middle Niger

### 2023-08-03 analysis/paper/paper.Rmd
Initial commit

---
title: "analysis"
author: "Nikolas Gestrich"
date: '2022-05-02'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(knitr.kable.NA = "")

# libraries
library(dplyr)
library(knitr)
library(patchwork)
library(ggplot2)
library(tidygraph)
library(ggraph)
library(tibble)
library(igraph)
library(tidyr)
library(sf)
```

## Data import
We are importing data on pottery decors from a variety of sources. All of the data represents phases of sites that are dated to have been occupied at some point in the 11th century AD.

XXX: put together the different periods. 
Update: or maybe not, so that we can name the nodes the same

```{r}
library(readr)
# import decor data
decors <- read_csv2(here::here("analysis/data/raw_data/decors_late.csv"))
# import information on sites
sites <- read_csv2(here::here("analysis/data/raw_data/sites_all.csv"))
```

show the data:

```{r fig.cap="decor data used for network construction"}
library(kableExtra)
kbl(decors) |> 
  kable_styling(font_size = 7, html_font = "Cambria") 
```

show the map points:

```{r fig.cap="location of sites in the analysis", warning=FALSE}
source(here::here("analysis/R/map plots.R"))

@@

sitemap <- ggplot() +
  map_bw_grid +
  geom_sf(data = sitepoints)

@@

We can see here that sites are geographically unequally distributed. This is due to research history as well as past cultural dynamics. Certainly, the IND and the areas bordering it (notiably the Lakes region) were strongly populated at the time, maybe more so than other areas. The cluster of sites around the Bandiagara escarpment is due to the extraordinarily detailed and long-term research of the University of Geneva's research teams. Blank spaces do not imply a lack of people, in this case people making and using pottery. Global politics and research priorities over the last 50 years have meant that these are areas we do not currently have excavated and dated ceramics for, though we know that sites exist, often in abundance. This means that, when we construct networks, we do so with a geographically unrepresentative sample, which leads to incomplete networks. Sindbaek has pointed out the difficulty in this situation: the actual network is not known, only its effects, from which the network can be partially reconstructed. Accordingly, this study should be seen as an exercise first in the reconstruction of a possible network that led to the data preserved in the archaeological record: a tiny sample of a tiny sample, influenced by an unknown amount of mostly unknowable pre- and post-depositional biases. Nevertheless, we will attempt to analyse the reconstructed network in order to formulate some hypotheses about the nature of innovation and spread in ceramic decors. Future generations of researchers are encouraged to improve upon this initial attempt.

# Construct network

```{r}
# import functions for network construction
source(here::here("analysis/R/edgelist functions.R"))

@@

# construct network, remove extraneous sites
decor_graph <- tbl_graph(sites, edgelist, directed = FALSE) %>%
  activate(nodes) %>%
  filter(!node_is_isolated())

decor_graph
```

# Plot network

```{r total network, fig.cap="Plot of 11th century AD network of decorative practices. The plot shows the very strongly connected structure of the network."}
ggraph(decor_graph) +
  coloured_edges +
  theme_graph()
```

## Network structure

High number of connections due to the construction of the network. This network is very much unsuitable for further network analysis, but it already allows us to see one major thing about pottery decor interactions: the networks in which individual decor types are propagated are overlapping. There are no two decor networks that cover one another entirely.

```{r total network sample, fig.cap="Partially overlapping networks of four decorative practices"}
decor_graph %>%
  activate(edges) %>%
  filter(name %in% c("braided_cord_roulette", "braided_strip_roulette", "paint", "straight_cord_mat")) %>%
  ggraph() +
  coloured_edges +
  theme_graph()
```

## Calculate difference between network layers

We do this using the `multinet` package.

```{r import to multinet}
library(multinet)
# reduce graph to types with over 5 edges
good_edges <- decor_graph %>%
  activate(edges) %>%
  as_tibble() %>%
  count(name) %>%
  filter(n > 5)

@@

```{r layer comparison, fig.cap="Jaccard similarity of nodes present in the networks of individual decorative practices"}
# perform layer compariosn using jaccard similarity between nodes
lc <- layer_comparison_ml(multi, method = "jaccard.actors")

@@

This shows us the jaccard similarity of the actors present on the different levels of the network. We can see that there are more similar and more dissimilar layers in the network.Very high are the similarities between folded strip roulette and braided cord roulette, two common decors. Also high is the similarity in common decorative techniques of channeling and applied decors. The fact that there are similar and dissimilar layers suggests that we might be able to identify groups in the network.

## Clustering

We use the Louvain clustering algorithm (Blondel et al 2008) to find communities in the network.

```{r community detection, hide = 'results', echo=FALSE}
# use louvain clustering method to assign clusters to nodes
decor_graph <- decor_graph %>%
  activate(nodes) %>%
  mutate(louvain = as.character(group_louvain()))

# calculate modularity of louvain clustering
modularity <- decor_graph %>%
  activate(nodes) %>%
  mutate(modularity = graph_modularity(group = as.factor(louvain))) %>%
  pull(modularity) %>%
  head(1)
```

The louvain clustering has a modularity of `r modularity` .

```{r plot-clusters, fig.cap="Louvain communities within the decor network"}
# set plot colours for communities
communitycolor <-
  setNames(
    c("tomato", "cornflowerblue", "darkgreen"),
    unique(V(decor_graph)$louvain)
  )

@@

```{r setting up colors and multilevel community, hide = 'results', echo=FALSE}
## in the final form this part will be much shorter because the colors are set elsewhere
# we need a vector of colors that is named after the value we want to be plotted (community, name of the decor)
communitycolor <-
  setNames(
    c("tomato", "cornflowerblue", "darkgreen"),
    unique(V(decor_graph)$louvain)
  )

# because decor is longer we need to create a palette with the proper amount of colors (the whole code is still not perfect, but i think it is the proper way that can still be optimized)
col_21 <- colorRampPalette(brewer.pal(8, "Set2"))(21)

# then we create a vector containing all decortypes
decortypes <- decor_graph %>%
  activate(edges) %>%
  as_data_frame(what = "edges") %>%
  select(name) %>%
  unique()

decortypes <- as.vector(decortypes[, 1])

# finally we make a named vector of the colors
decorcolor <- setNames(col_21, decortypes)
```

@@

Here, we employ the same clustering algorithm, which results in three communities with a modularity of 0.3. Community 1 contains 7 Sites in the Dogon Area. Community 2 is in the Niger bend and the IND, it also contains 7 sites. Both only include sites south of the Niger. Community 3 is the biggest and most widely distributed community. Reaching from Gao to Togu Missiri in the Ségu area it seems to mostly contain sites around the Niger river. The nodes of community 3 often connect with sites of community 1 and 2, while those communities are not connected strongly.

@@

```{r cluster multilevel, echo=FALSE}
nodes <- decor_graph %>%
  activate(nodes) %>%
  as_tibble()
hull1 <- nodes %>%
  filter(louvain == 1) %>%
  slice(chull(lon, lat)) %>%
  rename(x = lon, y = lat)

hull2 <- nodes %>%
  filter(louvain == 2) %>%
  slice(chull(lon, lat)) %>%
  rename(x = lon, y = lat)

hull3 <- nodes %>%
  filter(louvain == 3) %>%
  slice(chull(lon, lat)) %>%
  rename(x = lon, y = lat)

# override s2 use for geographic plotting
sf_use_s2(FALSE)

ggraph(decor_graph, x = lon, y = lat) +
  # plot on map background
  bw.dem.map +
  # set edge parameters
  geom_edge_fan(color = "gray", aes(edge_alpha = weight, edge_width = weight)) +
  scale_edge_width(range = c(0.2, 2)) +
  scale_edge_alpha(range = c(0.3, 0.6)) +
  # colour nodes according to cluster and label
  geom_node_point(aes(color = louvain)) +
  scale_color_manual(values = communitycolor) +
  geom_node_text(aes(label = label), col = "black", repel = TRUE) +
  # draw cluster hulls
  geom_polygon(data = hull1, aes(x = x, y = y), alpha = 0.5, fill = "cornflowerblue") +
  geom_polygon(data = hull2, aes(x = x, y = y), alpha = 0.5, fill = "tomato") +
  geom_polygon(data = hull3, aes(x = x, y = y), alpha = 0.5, fill = "darkgreen") +
  # set colour scales
  labs(color = "Community") +
  guides(color = guide_legend(override.aes = list(size = 0.5)), 
         alpha = "none", size = "none", edge_alpha = "none", 
         edge_width = guide_legend(override.aes = list(size = 0.3))) +
  ggtitle("Network with multilevel communities") +
  # set map extent
  coord_sf(xlim = c(-7, 2), ylim = c(13, 19), expand = FALSE) +
  theme_graph()
```

```{r relative edge weights and centralization by cluster}
# Relative edge weights (how important are the connections within the cluster)
rel_edge_weight <- vector("double", length(3))
for (i in 1:3) {
  rel_edge_weight[[i]] <- decor_graph %>%
    # filter community
    activate(nodes) %>%
    filter(louvain == i) %>%
    # relative edge weight (sum of weights / number of edges)
    activate(edges) %>%
    as.data.frame() %>%
    {
      sum(.$weight) / nrow(.)
    }
}

@@

# put together
tibble(rel_edge_weight, centralization) %>%
  mutate(cluster = 1:3) %>%
  mutate(rel_edge_weight = rel_edge_weight / 100) %>%
  rename("relative edge weight" = rel_edge_weight) %>%
  pivot_longer(!cluster, names_to = "key", values_to = "value") %>%
  # plot
  ggplot(aes(x = cluster, y = value, fill = key)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_grey() +
  theme(axis.text.x = element_text(angle = 90)) +
  theme_publish()
```

```{r edges by cluster, eval = FALSE}

edges.com1 <- decor_graph %>%
  activate(nodes) %>%
  filter(louvain == 1) %>%
  activate(edges) %>%
  as_data_frame(what = "edges") %>%
  group_by(name) %>%
  summarise(n = sum(weight)) %>%
  filter(n > 0) %>%
  mutate(community = 1)

edges.com2 <- decor_graph %>%
  activate(nodes) %>%
  filter(louvain == 2) %>%
  activate(edges) %>%
  as_data_frame(what = "edges") %>%
  group_by(name) %>%
  summarise(n = sum(weight)) %>%
  filter(n > 0) %>%
  mutate(community = 2)

edges.com3 <- decor_graph %>%
  activate(nodes) %>%
  filter(louvain == 3) %>%
  activate(edges) %>%
  as_data_frame(what = "edges") %>%
  group_by(name) %>%
  summarise(n = sum(weight)) %>%
  filter(n > 0) %>%
  mutate(community = 3)

# edges.comcrossing <- decor_graph %>%
# activate(edges) %>%
#  mutate(crossing = igraph::crossing(louvain,.)) %>%
#  filter(crossing == TRUE) %>%
#  as_data_frame(what = "edges") %>%
# group_by(name) %>%
#  summarise(n = sum(weight)) %>%
#  filter(n > 0) %>%
#    mutate(community = "crossing")

plyr::rbind.fill(
  edges.com1, edges.com2, edges.com3
  # , edges.comcrossing
) %>%
  filter(n != 0) %>%
  mutate(community = as.factor(community)) %>%
  ggplot(aes(x = reorder(name, -n, sum), y = n, fill = community)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("tomato", "cornflowerblue", "darkgreen", "yellow")) +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_x_discrete(name = "decor type") +
  scale_y_continuous(name = "sum of the weight") +
  ggtitle("decors by clusters")

str.1 <- edges.com1 %>%
  select(n) %>%
  sum()

str.2 <- edges.com2 %>%
  select(n) %>%
  sum()

str.3 <- edges.com3 %>%
  select(n) %>%
  sum()

com.edges <- c(str.1 / nrow(edges.com1), str.2 / nrow(edges.com2), str.3 / nrow(edges.com3)) / 10000

library(tibble)
cent.1 <- decor_graph %>%
  filter(louvain == 1) %>%
  activate(nodes) %>%
  centr_degree() %>%
  as.tibble() %>%
  select(centralization) %>%
  transmute(centralization.weight = (centralization / nrow(.))) %>%
  slice(1)

cent.2 <- decor_graph %>%
  activate(nodes) %>%
  filter(louvain == 2) %>%
  centr_degree() %>%
  as.tibble() %>%
  select(centralization) %>%
  transmute(centralization.weight = (centralization / nrow(.))) %>%
  slice(1)

cent.3 <- decor_graph %>%
  activate(nodes) %>%
  filter(louvain == 3) %>%
  centr_degree() %>%
  as.tibble() %>%
  select(centralization) %>%
  transmute(centralization.weight = (centralization / nrow(.))) %>%
  slice(1)

library(tidyr)
community.edges <- bind_rows(cent.1, cent.2, cent.3) %>%
  add_column("sum of com.edges" = com.edges) %>%
  gather(key) %>%
  add_column(community = c(1:3, 1:3))

ggplot(community.edges, aes(x = community, y = value, fill = key)) +
  geom_bar(stat = "identity", position = "dodge") +
  # scale_fill_manual(values = c("tomato", "cornflowerblue", "darkgreen"))+
  theme(axis.text.x = element_text(angle = 90))
```
If we look at the edges (divided by 10000 to make it comparable) and the centrality of sites in regard to the communities, we see that while the sites of community 3 are the most central ones in the network the weight of the edges within the community is relatively small. For community 1 and 2 it is vice versa even though community 2 has weaker edges within and is better connected to the whole network. The more isolated a community is from the network, the stronger is its connection within the community. 
--maybe the bar_plots of the most common decors of all clusters is enough? maybe the first barplot is enough? – maybe the network of every cluster is too much
If we take a closer look at the "internal" connections of community 1, we see that "braided_strip_roulette" is present at all sites. The other main decor "straight_cord_mat" on the other hand is missing on 2 sites: Oudalan and Prom. The two decors are spread across the whole area of the community. The only other decor that is common in the community is the "twisted_cord_roulette" which is also abundant across the whole research area. It makes up a rather large part of the assemblage in Oudalan and Prom but was not found at the sites of Sadia and Ambere Dougon.
The fact that SAO has a high degree of centrality shows that the location is not decisive for the connectedness within this community. This is underlined by the fact that the less central sites are distributed across the whole area. There is an ambivalence between a relative consistency across the whole area and the diversity between sites that are close to each other. One reason for this phenom can be the contact to other communities in the mountain area. Damassogou exemplifies this as the site contains the defining decors of community 1 and 2. But the difference between Oudalan and SAO that are geographically isolated suggest that there might be more reasons for this diversity.
Community 2 is less diverse with "braided_cord_roulette" as the only outstanding decor type. The weaker connections of the sites Kokolo and Damassogou are based on data bias (the proportion is counted in regard to all sherds not only the decorated ones), the decor was probably equally abundant there. Other decors spread across the community are "folded_strip_roulette" and "dragged_comb". The first one is missing in KAN 4 and TOD H while the second one is missing at Kokolo and Damassogou. The other indicator for the homogeneous character of the community is the nearly equal degree of centrality of every site.
As mentioned above community 3 is more diverse but also has weaker connections. Instead of one defining decor there are several ones but none is abundant across the whole community. The spatial distribution differs between the decors. While "paint" and "twisted_cord_roulette" appear from Gao to TOG 2, "folded_strip_roulette", "channel" and "incised" are located in the western part of the community's area. Essouk and TMD have the highest degree of centrality. Even though it is located at the margin of the research area, its character as a "market"-site explains the amount of links it has. The role of TMD for the network is elaborated elsewhere (!).

```{r crossings}
decor_graph %>%
  activate(edges) %>%

@@

  geom_edge_fan(aes(color = name, edge_width = weight, edge_alpha = weight)) +
  # scale_edge_color_manual(values = decorcolor) +
  scale_edge_width(range = c(0.2, 2)) +
  geom_node_point(aes(size = centrality_eigen(weights = weight), color = louvain), show.legend = FALSE) +
  geom_node_text(aes(label = label), col = "black", repel = TRUE) +
  scale_color_manual(values = communitycolor)+

  theme_graph() +
  theme_community

```

```{r crossings between communities, eval = FALSE}

## this plot is hard to read but contains a lot of information, it shows only edges that connect the 2 communities. the color of the nodes shows their community while the size shows the betweenness centrality. This way we can show which sites are on the at the "frontier" of the communities that have strong connections to the other community. only edgest that are crossings between communities are shown, the widest ones are those that are those with an higher betweenness. in this way the graph shows the relation between the communities and which decor is shared/crossing community boundaries

nodes.c1 <- decor.graph %>%
  activate(nodes) %>%
  filter(communityl == 1) %>%
  as_data_frame(what = "vertices") %>%
  pull(id)

nodes.c2 <- decor.graph %>%
  activate(nodes) %>%
  filter(communityl == 2) %>%
  as_data_frame(what = "vertices") %>%
  pull(id)

nodes.c3 <- decor.graph %>%
  activate(nodes) %>%
  filter(communityl == 3) %>%
  as_data_frame(what = "vertices") %>%
  pull(id)

c1_3 <- decor.graph %>%
  activate(edges) %>%
  mutate(crossing = crossing(cl, .)) %>%
  filter(crossing == TRUE) %>%
  as_data_frame(what = "edges") %>%
  filter(from %in% c(nodes.c3) | to %in% c(nodes.c3)) %>%
  filter(from %in% c(nodes.c1) | to %in% c(nodes.c1)) %>%
  group_by(name) %>%
  summarise(n = sum(weight)) %>%
  filter(n > 0) %>%
  mutate(crossing = "c1_3")

c2_3 <- decor.graph %>%
  activate(edges) %>%
  mutate(crossing = crossing(cl, .)) %>%
  filter(crossing == TRUE) %>%
  as_data_frame(what = "edges") %>%
  filter(from %in% c(nodes.c3) | to %in% c(nodes.c3)) %>%
  filter(from %in% c(nodes.c2) | to %in% c(nodes.c2)) %>%
  group_by(name) %>%
  summarise(n = sum(weight)) %>%
  filter(n > 0) %>%
  mutate(crossing = "c2_3")

>>>>

---
title: "Pottery decor as networks on the Middle Niger"
author:
  - Nikolas Gestrich:
      email: gestrich@uni-frankfurt.de
      institute: [FI]
      correspondence: true
  - Juan-Marco Puerta-Schardt:
      email: jmpuertaschardt@outlook.de
      institute: [GU]
      correspondence: false
institute:
  - FI: Frobenius Institute
  - GU: Goethe University
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      #pandoc_args:
      #- --lua-filter=../templates/scholarly-metadata.lua
      #- --lua-filter=../templates/author-info-blocks.lua
      #- --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The following code chunk defines some general settings how code chunks should behave. -->

```{r knitr-setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)
```

```{r libraries, echo = FALSE}
# libraries
library(dplyr)
library(knitr)
library(patchwork)
library(ggplot2)
library(tidygraph)
library(ggraph)
library(tibble)
library(igraph)
library(tidyr)
library(sf)
```

<!-- The actual document text starts here: -->

# Introduction

For much of the history of the archaeological discipline, ceramics have been used as markers of similarity and difference.In many areas and periods of the past, they are the most ubiquitous archaeological find category, and have thus contributed greatly to the recognition of archaeological artefact groups, or archaeological cultures. It is not rare to see find complexes and periods named after particular ceramic types. Though the current scholarly consensus is that archaeological cultures do not represent social groups as experienced by past humans, and that pots do not equal people, a large amount of work on archaeological ceramics is engaged in grouping ceramics and distinguishing groups from one another. There are a number of approaches for doing so and the rationales behind them vary. The basis for the creation of archaeological cultures, especially current in works of the first two thirds of the twentieth century, is the 'traditional' typological approach to ceramics and other artefacts. This approach is nowadays most used in the best-explored regions with the longest research traditions, and groups ceramics mainly according to outward appearances, such as shape and surface treatment. The distribution, both spatially and temporally, of such types then becomes the subject of research, which subsequent studies can augment through the recognition of sub-types and variants. A different approach takes as its basis the sequence of manufacture of artefacts, the *chaîne opératoire*. This approach proposes that the manufacture of an artefact involves a series of choices constrained by certain outside factors. These choices are seen as being cultural, so that a similar set of choices under similar constraints are markers of shared cultural knowledge. Here, the archaeological research goal also lies in the identification of spheres of cultural commonality which can be distinguished from one another.

In the archaeology of West Africa, and especially in the wider areas surrounding the upper and middle reaches of the Niger, there has been a strong research focus on ceramics. From the 1st millennium BC onwards, they occur in extremely high numbers on archaeological sites, and their appearance, especially their surface treatment, is highly variable. In contrast to the Great Lakes region of Eastern Africa, for example, the number of different roulette decorations, painted, incised, and applied decors, but also the variety in form, is very high. Consequently, there has been considerable emphasis on ceramics in archaeological and also in ethnoarchaeological work. While much of the ethnoarchaeological work has been based on exploring *chaînes opèratoires*, the archaeological work has mainly relied on attribute-based analyses (the system employed by S.K. McIntosh at Jenné-jeno has been very influential in subsequent work). In a region where large archaeological sites abound and researchers are relatively thin on the ground, the main hope of the analyses usually is to create chrono-stratigraphical sequences in order to be able to give approximate periods to surface assemblages. Despite all the variability, it has been found that a great number of ceramic attributes are not particularly time-sensitive. On the contrary, most sites appear to have diverse assemblages in which the percentages of individual attributes vary, but persist over several centuries, sometimes millennia.

Some researchers have used the attribute-based approach to define types, statistically derived for the most part, which typify the ceramics of a certain place of time (MacDonald's Facies, McIntosh's timeline). The spatiotemporal distribution of individual attributes has also been the subject of some important work, such as individual rouletted decors (Livingstone-Smith) or base-shaping techniques (Huysecom). The most comprehensive treatment so far is in the work of Anne Mayor, combining both shaping techniques and decorative practices from an archaeological and ethnoarchaeological point of view. Focused on the Bandiagara highlands in central Mali, Mayor uses her data to construct an impressive overview of the development of different techniques and the establishment of ceramic traditions, which she closely relates to ethno-linguistic groups. Mayor's work builds on a large body of data and decades of ethnoarchaeological fieldwork, and there is little to add to her conclusions. Here, however, we wish to show that an almost identical dataset as available to Mayor can generate quite a different perspective when analysed as a network, and when connectivity is emphasised over boundaries.

# Archaeological ceramics as networks

In recent years, the use of network analysis in archaeology has been growing significantly. A large number of these studies focus on ceramics, but they are geographically mainly limited to North America (e.g. Hart 2012, Chiu 2019, Barker and Young 2017, Crabtree 2015, Mills et al. 2013, Peeples and Haas 2013) and the Mediterranean (e.g. Sternini 2019, Iacono 2016, Brughmans and Poblome 2016 torrence & Swadling 2008, de Groot 2019). The archaeological records of these regions are amongst the best studied in the world, with fine-grained pottery typologies and extensive excavations. To our knowledge, this is the first application of a network approach to archaeological material from sub-Saharan Africa, where the conditions are somewhat different.

Ethnographic studies of pottery manufacturing in Africa suggest that the way that pottery is made depends largely on systems of learning and teaching the craft (Gosselain), on local market consumer choices (Dietler & Herbich), or on marriage patterns amongst manufacturing groups (ibid.). In short, the processes underlying the propagation of technical practices through time and across geographical space are dependent on communication and, more specifically, knowledge exchange. Such exchange occurs between humans and we can therefore understand similarities in the *chaine opératoire* as evidence that knowledge was shared. However, different elements of the pottery making process are thought to be reflective of different kinds of social interaction. Gosselain (REF), whose research in the years between XX and XX explored this question, concludes that, on the whole, forming techniques are more stable, reflective of durable cultural ties, sometimes even ethnic groups. On the other end of the spectrum, decorations are thought to be influenced by far more ephemeral contact, in which seeing someone else's pot might be enough to change a potter's production, as could indirect contact via the wishes of customers.

SNA shifts the focus "from spatial relations to social relations" (cf. Knappett 2011)
Mills et al. (2015) have been able to compare similarity in wares across sites, but this cannot be done here. They also highlight Knappet's (2011) idea of different scales in archaeological networks. Because of the specific use case scenario here, we are conflating two fo his scales. He sees the community of practice as a meso-scale effect, across maybe a few households. However, we can see in the archaeological ceramic data from Mali that communities of practice in individual stages of the *chaîne operatoire* is often much larger. Thus, certain parts of the pottery manufacturing sequence are distributed across larger or smaller networks, but rarely remain at the household level. Also, we do not really have the luxury of being able to pinpoint certain wares, since typology is not as established.

We need to be aware of the fact that we are conflating consumption with production here. This is why we will need to threshold the networks, to make sure that a site represents the significant potting communities within its area.

# Methods

```{r import 11c decor data, echo = FALSE}
library(readr)
# import decor data
decors <- read_csv2(here::here("analysis/data/raw_data/decors_late.csv"))
# import information on sites
sites <- read_csv2(here::here("analysis/data/raw_data/sites_all.csv"))
```

```{r datatable, fig.cap="decor data used for network construction"}
library(kableExtra)
kable(decors, digits = 1, "html") %>% 
  kable_styling("condensed") %>% 
  save_kable(here::here("analysis/figures/datatable.html"))
```

In this paper, we use pottery decor data from contexts occupied in the 11th century AD to attempt a reconstruction of networks through which pottery decorative practices were shared. The data is in a common format of percentages of prevalence in the assemblage (see \@ref(tab:datatable) This particular time slice, which represents a cumulative view of a great many individual interactions, will then be compared to earlier contexts in the same wider area, in order to demonstrate the diachronic possibilities of the approach.The sites we have been able to include in the network model are shown in \@ref(fig:sitemap). We can see here that sites are geographically unequally distributed. This is due to research history as well as past cultural dynamics. Certainly, the IND and the areas bordering it (notably the Lakes region) were strongly populated at the time, maybe more so than other areas. The cluster of sites around the Bandiagara escarpment is due to the extraordinarily detailed and long-term research of the University of Geneva's research teams. Blank spaces do not imply a lack of people, in this case people making and using pottery. Global politics and research priorities over the last 50 years have meant that these are areas we do not currently have excavated and dated ceramics for, though we know that sites exist, often in abundance. This means that, when we construct networks, we do so with a geographically unrepresentative sample, which leads to incomplete networks. Sindbaek has pointed out the difficulty in this situation: the actual network is not known, only its effects, from which the network can be partially reconstructed. Accordingly, this study should be seen as an exercise first in the reconstruction of a possible network that led to the data preserved in the archaeological record: a tiny sample of a tiny sample, influenced by an unknown amount of mostly unknowable pre- and post-depositional biases. Nevertheless, we will attempt to analyse the reconstructed network in order to formulate some hypotheses about the nature of innovation and spread in ceramic decors. Future generations of researchers are encouraged to improve upon this initial attempt.

```{r sitemap, fig.cap="location of sites in the analysis", echo = FALSE}
source(here::here("analysis/R/map plots.R"))

@@

sitemap <- ggplot()+
  map_bw_grid +
  geom_sf(data = sitepoints)

@@

The network is constructed from standard archaeological frequency data, in which the proportions of decorative practices are represented across the selected site phases (REF FIG XXX). The decorative practices encode both the tool used and the action performed, for instance a twisted cord (tool) roulette (action). This data is transformed into a network by deciding that every decor that is shared between two sites leads to a link between these sites. The link is weighted by multiplying the proportion at one site with the proportion at the other. This leads to a network in which the nodes (sites) can be connected in more than one manner, known as a multiplex or multilayer network. As is shown in \@ref(fig:total-network), this leads to a very tightly connected network, known as a "hairball". However, the multiplex nature of the network allows us to make an important initial observations on its structure: the networks in which individual decor types are propagated are overlapping. There are no two decor networks that cover one another entirely. This emerges more clearly in \@ref(fig:total-network-sample), when the number of decorative practices is reduced. This indicates that the adoption of decorative techniques ...

```{r make decor network, echo = FALSE}
# import functions for network construction
source(here::here("analysis/R/edgelist functions.R"))

@@

# construct network, remove extraneous sites
decor_graph <- tbl_graph(sites, edgelist, directed = FALSE) %>%
  activate(nodes) %>%
  filter(!node_is_isolated())
```

```{r total-network, fig.cap="Plot of 11th century AD network of decorative practices. The plot shows the very strongly connected structure of the network."}
ggraph(decor_graph, layout = "stress") +
  coloured_edges +
  theme_graph()
```

```{r total-network-sample, fig.cap="Partially overlapping networks of four decorative practices", echo = FALSE}
decor_graph %>%
  activate(edges) %>%
  filter(name %in% c("braided_cord_roulette", "braided_strip_roulette", "paint", "straight_cord_mat")) %>%
  ggraph(layout = "stress") +
  coloured_edges +
  theme_graph()
```

Using the `multinet` package, the statistical distance (Jaccard similarity) between the different decors' networks were calculated based on the nodes they contained. We see in \@ref(fig:layer-comparison) that some networks are more similar to one another than others. This leads us to ask whether there might be groups of decors that were shared within similar networks, and thus, whether there might be groups of sites predominantly sharing specific decors.

```{r import to multinet, echo = FALSE}
library(multinet)
# reduce graph to types with over 5 edges
good_edges <- decor_graph %>%
  activate(edges) %>%
  as_tibble() %>%
  count(name) %>%
  filter(n > 5)

@@

```{r layer-comparison, fig.cap="Jaccard similarity of nodes present in the networks of individual decorative practices"}
# perform layer compariosn using jaccard similarity between nodes
lc <- layer_comparison_ml(multi, method = "jaccard.actors")

@@

# Clustering

We use the Louvain clustering algorithm (Blondel et al 2008) to find communities in the network.
```{r community detection, hide = 'results', echo=FALSE}
# use louvain clustering method to assign clusters to nodes
decor_graph <- decor_graph %>%
  activate(nodes) %>%
  mutate(louvain = as.character(group_louvain()))

# calculate modularity of louvain clustering
modularity <- decor_graph %>%
  activate(nodes) %>%
  mutate(modularity = graph_modularity(group = as.factor(louvain))) %>%
  pull(modularity) %>%
  head(1)
```
The louvain clustering has a modularity of `r modularity` .

```{r plot-clusters, fig.cap="Louvain communities within the decor network", echo = FALSE}
# set plot colours for communities
communitycolor <-
  setNames(
    c("tomato", "cornflowerblue", "darkgreen"),
    unique(V(decor_graph)$louvain)
  )

@@

@@

Here, we employ the same clustering algorithm, which results in three communities with a modularity of 0.3. See \@ref(fig:plot-clusters) Community 1 contains 7 Sites in the Dogon Area. Community 2 is in the Niger bend and the IND, it also contains 7 sites. Both only include sites south of the Niger. Community 3 is the biggest and most widely distributed community. Reaching from Gao to Togu Missiri in the Ségu area it seems to mostly contain sites around the Niger river. The nodes of community 3 often connect with sites of community 1 and 2, while those communities are not connected strongly.

@@

```{r cluster-comparison, fig.cap="Relative edge weights and centralization by cluster"}
# Relative edge weights (how important are the connections within the cluster)
rel_edge_weight <- vector("double", length(3))
for (i in 1:3) {
  rel_edge_weight[[i]] <- decor_graph %>%
    # filter community
    activate(nodes) %>%
    filter(louvain == i) %>%
    # relative edge weight (sum of weights / number of edges)
    activate(edges) %>%
    as.data.frame() %>%
    {
      sum(.$weight) / nrow(.)
    }
}

@@

# put together
tibble(rel_edge_weight, centralization) %>%
  mutate(cluster = 1:3) %>%
  mutate(rel_edge_weight = rel_edge_weight / 100) %>%
  rename("relative edge weight" = rel_edge_weight) %>%
  pivot_longer(!cluster, names_to = "key", values_to = "value") %>%

  # plot
  ggplot(aes(x = cluster, y = value, fill = key)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_grey() +
  theme(axis.text.x = element_text(angle = 90)) +
  theme_publish()
```

If we look at the edges (divided by 10000 to make it comparable) and the centrality of sites in regard to the communities, we see that while the sites of community 3 are the most central ones in the network the weight of the edges within the community is relatively small. For community 1 and 2 it is vice versa even though community 2 has weaker edges within and is better connected to the whole network. The more isolated a community is from the network, the stronger is its connection within the community. 
--maybe the bar_plots of the most common decors of all clusters is enough? maybe the first barplot is enough? – maybe the network of every cluster is too much

 see \@ref(fig:cluster-comparison)
If we take a closer look at the "internal" connections of community 1, we see that "braided_strip_roulette" is present at all sites. The other main decor "straight_cord_mat" on the other hand is missing on 2 sites: Oudalan and Prom. The two decors are spread across the whole area of the community. The only other decor that is common in the community is the "twisted_cord_roulette" which is also abundant across the whole research area. It makes up a rather large part of the assemblage in Oudalan and Prom but was not found at the sites of Sadia and Ambere Dougon.
The fact that SAO has a high degree of centrality shows that the location is not decisive for the connectedness within this community. This is underlined by the fact that the less central sites are distributed across the whole area. There is an ambivalence between a relative consistency across the whole area and the diversity between sites that are close to each other. One reason for this phenom can be the contact to other communities in the mountain area. Damassogou exemplifies this as the site contains the defining decors of community 1 and 2. But the difference between Oudalan and SAO that are geographically isolated suggest that there might be more reasons for this diversity.
Community 2 is less diverse with "braided_cord_roulette" as the only outstanding decor type. The weaker connections of the sites Kokolo and Damassogou are based on data bias (the proportion is counted in regard to all sherds not only the decorated ones), the decor was probably equally abundant there. Other decors spread across the community are "folded_strip_roulette" and "dragged_comb". The first one is missing in KAN 4 and TOD H while the second one is missing at Kokolo and Damassogou. The other indicator for the homogeneous character of the community is the nearly equal degree of centrality of every site.
As mentioned above community 3 is more diverse but also has weaker connections. Instead of one defining decor there are several ones but none is abundant across the whole community. The spatial distribution differs between the decors. While "paint" and "twisted_cord_roulette" appear from Gao to TOG 2, "folded_strip_roulette", "channel" and "incised" are located in the western part of the community's area. Essouk and TMD have the highest degree of centrality. Even though it is located at the margin of the research area, its character as a "market"-site explains the amount of links it has. The role of TMD for the network is elaborated elsewhere (!).

```{r crossings, fig.cap="Network of edges which cross between clusters"}
decor_graph %>%
  activate(edges) %>%

@@

  geom_edge_fan(aes(color = name, edge_width = weight, edge_alpha = weight)) +
  scale_edge_width(range = c(0.2, 2)) +
  geom_node_point(aes(size = centrality_degree(weights = weight), color = louvain), show.legend = FALSE) +
  geom_node_text(aes(label = label), col = "black", repel = TRUE) +
  scale_color_manual(values = communitycolor)+
  theme_graph() +
  theme_community
```

Not only the connections within but also those between the communities should be examined. Therefore, only the edges that "cross" the two communities will be considered. While there is no obvious "broker"-site, community 3 could be described as a "broker"-community. There are few strong crossings between community 1 and 2. Even though most decors that cross communities connect all three, those are mostly decors that are most abundant at the sites of community 3. See \@ref(fig:crossings)
